---
title: "Mini Data Analysis Milestone 2"
output: github_document
---

*To complete this milestone, you can either edit [this `.rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are commented out with `<!--- start your work here--->`. When you are done, make sure to knit to an `.md` file by changing the output in the YAML header to `github_document`, before submitting a tagged release on canvas.*

# Welcome to the rest of your mini data analysis project!

In Milestone 1, you explored your data. and came up with research questions. This time, we will finish up our mini data analysis and obtain results for your data by: 

- Making summary tables and graphs 
- Manipulating special data types in R: factors and/or dates and times.
-   Fitting a model object to your data, and extract a result.
-   Reading and writing data as separate files.

We will also explore more in depth the concept of *tidy data.* 

**NOTE**: The main purpose of the mini data analysis is to integrate what you learn in class in an analysis. Although each milestone provides a framework for you to conduct your analysis, it's possible that you might find the instructions too rigid for your data set. If this is the case, you may deviate from the instructions -- just make sure you're demonstrating a wide range of tools and techniques taught in this class.

# Instructions

**To complete this milestone**, edit [this very `.Rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work here--->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to your mini-analysis GitHub repository, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 50 points: 45 for your analysis, and 5 for overall reproducibility, cleanliness, and coherence of the Github submission. 

**Research Questions**: In Milestone 1, you chose two research questions to focus on. Wherever realistic, your work in this milestone should relate to these research questions whenever we ask for justification behind your work. In the case that some tasks in this milestone don't align well with one of your research questions, feel free to discuss your results in the context of a different research question.

# Learning Objectives

By the end of this milestone, you should:

-   Understand what *tidy* data is, and how to create it using `tidyr`.
-   Generate a reproducible and clear report using R Markdown.
-   Manipulating special data types in R: factors and/or dates and times.
-   Fitting a model object to your data, and extract a result.
-   Reading and writing data as separate files.

# Setup

Begin by loading your data and the tidyverse package below:

```{r, message = FALSE}
library(datateachr) # <- might contain the data you picked!
library(tidyverse)
```

# Task 1: Process and summarize your data 

From milestone 1, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions. 

### 1.1 (1 point) 

First, write out the 4 research questions you defined in milestone 1 were. This will guide your work through milestone 2:

<!-------------------------- Start your work below ---------------------------->
1. *How does diameter of the tree change if there is or is not a root barrier installed?*
2. *Does the number of trees in each neigborhood differ?*
3. *Do trees planted in 1995 and 2015 have a large difference in height_range_id?*
4. *What is the distribution of tree diameter in each neighborhood?*
<!----------------------------------------------------------------------------->

Here, we will investigate your data using various data manipulation and graphing functions.

### 1.2 (8 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question.

Ensure that the output of each operation is printed!

Also make sure that you're using dplyr and ggplot2 rather than base R. Outside of this project, you may find that you prefer using base R functions for certain tasks, and that's just fine! But part of this project is for you to practice the tools we learned in class, which is dplyr and ggplot2.

**Summarizing:**

1.  Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2.  Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3.  Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4. Compute the proportion and counts in each category of one categorical variable across the groups of another categorical variable from your data. Do not use the function `table()`!

**Graphing:**

6. Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read.
7. Make a graph where it makes sense to customize the alpha transparency.

Using variables and/or tables you made in one of the "Summarizing" tasks: 

8. Create a graph that has at least two geom layers. 
9. Create 3 histograms, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!

<!------------------------- Start your work below ----------------------------->
### Task 1.2

**Question 1: How does diameter of the tree change if there is or is not a root barrier installed?**

Summarizing Task 1: This task helps determine different measure of diameter based on whether or not
there is a root barrier installed. This will allow me to see if the diameter differs in terms of multiple statistics or just one.
```{r}
mean_diam_barr<- vancouver_trees %>% group_by(root_barrier) %>% summarise(mean_diameter = mean(diameter))
mean_diam_barr
range_diam_barr<- vancouver_trees %>% group_by(root_barrier) %>% reframe(range_diameter = range(diameter), .groups = 'drop')
range_diam_barr
med_diam_barr<- vancouver_trees %>% group_by(root_barrier) %>% summarise(med_diameter = median(diameter))
med_diam_barr
max_diam_barr<- vancouver_trees %>% group_by(root_barrier) %>% summarise(max_diameter = max(diameter))
max_diam_barr
min_diam_barr<- vancouver_trees %>% group_by(root_barrier) %>% summarise(min_diameter = min(diameter))
min_diam_barr
```

Graphing Task 2: I created a bar graph to show the difference in mean diameter when comparing whether a tree has a root barrier or not. The bar graph also includes labels on each bar, which denote the diameter.
```{r}
ggplot(mean_diam_barr, aes(root_barrier, mean_diameter)) + geom_col() + geom_text(aes(label = mean_diameter)) + xlab("Presence of Root Barrier") + ylab("Mean Diameter") + ggtitle("Mean Diameter Based on Presence of Root Barrier")
```


**Question 2: Does the number of trees in each neigborhood differ?**

Summarizing Task 2: I chose this task because it would allow me to see how many trees have been planted in each neighborhood.
```{r}
trees_by_neigh <- vancouver_trees %>% group_by(neighbourhood_name) %>% summarise(n = n())
trees_by_neigh
```

Graphing Task 2: I chose this task because it will easily allow me to see how many trees are in each neighbourhood in Vancouver. The text at the bar is helpful to see the true value when two neighborhoods are similar in number.
```{r}
ggplot(trees_by_neigh, aes(neighbourhood_name, n)) + geom_col() + geom_text(aes(label = n)) + xlab("Neighbourhood") + ylab("Number of Trees") + ggtitle("Number of Trees in Vancouver Neighbourhoods") + theme(axis.text.x = element_text(angle = 90)) 
```


**Question 3: How many small, medium, and large trees are there in each neighborhood?**

(I changed this question from Milestone 1, the question from Milestone 1 is specified in the task above. I changed this question because I wanted to do Task 3 with my dataset but I wouldn't be able to do it with the original question.)

Summarizing Task 3: I chose this task to see what the distribution of tree size is within a neighbourhood and if certain neighborhoods have smaller or larger trees based on diameter.
```{r}
vancouver_trees3 <- vancouver_trees %>% filter(na.rm = TRUE) %>% group_by(neighbourhood_name) %>% mutate(diameter_cat = cut(diameter,
breaks = c(0, 14, 19, 20), labels = c("small", "medium", "large"), include.lowest = TRUE)) 
```

Graphing Task 3: I chose this task because I wanted to see which category has the most trees, and the spread of diameter in each category. I did a google search to see what typical small, medium, and large trees are sized at. I like this way of looking at the data because I am able to see the tree in the data set that weren't considered small, medium, or large.

I think the best histogram is the one with 100 bins because it is more clear if there is a range of diameters within the categories small, medium and large.

```{r}
ggplot(vancouver_trees3, aes(diameter, after_stat(density) )) +
   geom_histogram(bins = 30) + facet_wrap(('diameter_cat')) + scale_x_continuous(name="Diameter", limits=c(0, 30)) 

ggplot(vancouver_trees3, aes(diameter, after_stat(density) )) +
   geom_histogram(bins = 5) + facet_wrap(('diameter_cat')) + scale_x_continuous(name="Diameter", limits=c(0, 30))

ggplot(vancouver_trees3, aes(diameter, after_stat(density) )) +
   geom_histogram(bins = 100) + facet_wrap(('diameter_cat')) + scale_x_continuous(name="Diameter", limits=c(0, 30))
```


**Question 4: How many trees of the Alba species were planted every year?**

(I changed this question from Milestone 1, the question from Milestone 1 is specified in the task above. I changed this question because I wanted to make a graph using a geom_path and this option made the most choice, to see how a certain species may be planted over the years and if there is a visible trend.)

Summarizing Task 4: I chose this task to first count the number of trees in each species which would have to be done in order to further figure out and plot how many trees were planted per year.
```{r}
vancouver_trees %>% group_by(species_name) %>% summarise(n = n())
```

Graphing Task 4: I chose this task because I wanted to see how the trend in how many trees are planted per year changes. I think it may have been interesting to see how many trees were planted in general per year and to see if that number has grown to counteract the effects of climate change.
```{r}
alba_per_yr <- vancouver_trees %>% filter(species_name == "ALBA") %>% group_by(date_planted) %>% summarise(n = n())

ggplot(alba_per_yr, aes(date_planted, n )) + scale_x_continuous(name="Year") + scale_y_continuous(name="Number of Trees Planted", limits = c(0,3)) + geom_point() + geom_path()
```
<!----------------------------------------------------------------------------->

### 1.3 (2 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!------------------------- Write your answer here ---------------------------->

I am much closer to answering my research questions. I changed some of my questions to try to complete the tasks from the lists before I realized I could have kept my questions the same, but gone about the tasks differently. So I did refine the questions that I had based on the tasks. If I didn't change the questions before, I probably would have refined the questions at this point. I think seeing the trend per year for how many Alba trees were grown was interesting. I think seeing how many trees are planted per year in general will be more interesting. Also, it was specifed to use dplyr and ggplot for Task 1, so I hope to use the tidyverse data to clean up the data. One thing I can do is separate the date_planted column into day month and year, to get a more accurate version of the number of trees planted per year. Also, it was interesting to see that having a root barrier may be causing a smaller diameter of the tree on average. For this question, I don't think I would want to refine it. Using density plots are a bit harder for this dataset as there are quite variable values entered for some columns. Possibly filtering the data further will allow for more informative visuals.

<!----------------------------------------------------------------------------->

# Task 2: Tidy your data 

In this task, we will do several exercises to reshape our data. The goal here is to understand how to do this reshaping with the `tidyr` package.

A reminder of the definition of *tidy* data:

-   Each row is an **observation**
-   Each column is a **variable**
-   Each cell is a **value**

### 2.1 (2 points)

Based on the definition above, can you identify if your data is tidy or untidy? Go through all your columns, or if you have \>8 variables, just pick 8, and explain whether the data is untidy or tidy.

<!--------------------------- Start your work below --------------------------->
date_planted: This isn't as tidy as can be for my needs because it is hard to use year as a variable.
species_name: I would say this is tidy because it is separate from the genus_name column, and it would allow the user to filter for a specific species or genus if needed.
tree_id: This seems tidy as each observation is given a tree id, which makes each row an observation.
root_barrier: This is tidy because it is a yes or no column containing a single variable.
std_street: This entry could be split into three columns to divide the cardinal direction, street name, and the type of street it is, whether its a Avenue, Street, or Road etc.
neighbourhood_name: This seems tidy because there is consistency between how each neighbourhood is specified as each neighbourhood has its own string.
curb: This seems tidy enough because there is a Yes or No only for this variable/column.
cultivar_name: This may not be tidy because the common_name also includes this data. The data in common_name can probably be split to avoid the double presence of the cultivar_name.

Overall, the dataset is tidy, but it is untidy in some cases depending on the research question. Depending on the user's needs, they can make the data more tidy if needed. Each row is an observation as there is a tree id for each entry. Each column is a variable, and each cell has value unless it is specified as NA where there is no record of that variable for that observation.
<!----------------------------------------------------------------------------->

### 2.2 (4 points)

Now, if your data is tidy, untidy it! Then, tidy it back to it's original state.

If your data is untidy, then tidy it! Then, untidy it back to it's original state.

Be sure to explain your reasoning for this task. Show us the "before" and "after".

<!--------------------------- Start your work below --------------------------->
I want to make the data more tidy by separating the date_planted column into day_planted, month_planted, and year_planted. This will be useful for me because I would like to see how the number of trees planted per year changes, and this was doable but not as straightforward as when the columns for the date are separated out and year is more accessible.
```{r}
vancouver_trees_datesep <- vancouver_trees %>% separate(col = date_planted, into = c('year_planted', 'month_planted', 'day_planted'))
vancouver_trees_datesep
```
Now, I can put the componenets of the date back together.
```{r}
vancouver_trees_datetog <- vancouver_trees_datesep %>% unite(col = date_planted, c(year_planted, month_planted, day_planted), sep = "-")
vancouver_trees_datetog
```

<!----------------------------------------------------------------------------->

### 2.3 (4 points)

Now, you should be more familiar with your data, and also have made progress in answering your research questions. Based on your interest, and your analyses, pick 2 of the 4 research questions to continue your analysis in the remaining tasks:

<!-------------------------- Start your work below ---------------------------->

1.  How has the number of trees planted changed per year per neighbourhood over the years that this data has been collected?
2.  How many different genuses of trees are there in each neighbourhood?

<!----------------------------------------------------------------------------->

Explain your decision for choosing the above two research questions.

<!--------------------------- Start your work below --------------------------->
1. For this question, it stems from my curiosity of trends in tree planting and its benefit for climate change. Also, I think it would be interesting to see if there are more trees planted in different neighbourhoods or if each neighbourhood plants the same amount of trees. If one neighbourhood already has a lot of trees, this visual will show that that neighbourhood already has enough trees planted.
2. I picked this question because I am curious to see how the diversity of trees in each neighbourhood may vary.For my inital question, I graphed how many trees were in each neighbourhood, but determining the diversity is more indicative of diversity of environments.
<!----------------------------------------------------------------------------->

Now, try to choose a version of your data that you think will be appropriate to answer these 2 questions. Use between 4 and 8 functions that we've covered so far (i.e. by filtering, cleaning, tidy'ing, dropping irrelevant columns, etc.).

(If it makes more sense, then you can make/pick two versions of your data, one for each research question.) 

<!--------------------------- Start your work below --------------------------->
Data Set for Question 1

```{r}
vancouver_trees_q1 <- vancouver_trees %>% separate(col = date_planted, into = c('year_planted', 'month_planted', 'day_planted')) %>% group_by(neighbourhood_name, year_planted) %>% summarise(n = n())
vancouver_trees_q1
```

Data Set for Question 2

```{r}
vancouver_trees_q2 <- vancouver_trees %>% group_by(neighbourhood_name) %>% summarise(distinct_genus = n_distinct(genus_name))
vancouver_trees_q2
```

# Task 3: Modelling

## 3.0 (no points)

Pick a research question from 1.2, and pick a variable of interest (we'll call it "Y") that's relevant to the research question. Indicate these.

<!-------------------------- Start your work below ---------------------------->

**Research Question**: How does diameter of the tree change if there is or is not a root barrier installed?

**Variable of interest**: diameter

<!----------------------------------------------------------------------------->

## 3.1 (3 points)

Fit a model or run a hypothesis test that provides insight on this variable with respect to the research question. Store the model object as a variable, and print its output to screen. We'll omit having to justify your choice, because we don't expect you to know about model specifics in STAT 545.

-   **Note**: It's OK if you don't know how these models/tests work. Here are some examples of things you can do here, but the sky's the limit.

    -   You could fit a model that makes predictions on Y using another variable, by using the `lm()` function.
    -   You could test whether the mean of Y equals 0 using `t.test()`, or maybe the mean across two groups are different using `t.test()`, or maybe the mean across multiple groups are different using `anova()` (you may have to pivot your data for the latter two).
    -   You could use `lm()` to test for significance of regression coefficients.

<!-------------------------- Start your work below ---------------------------->
```{r}
model_diam <- lm(diameter ~ root_barrier, vancouver_trees)
print(model_diam)
```
<!----------------------------------------------------------------------------->

## 3.2 (3 points)

Produce something relevant from your fitted model: either predictions on Y, or a single value like a regression coefficient or a p-value.

-   Be sure to indicate in writing what you chose to produce.
-   Your code should either output a tibble (in which case you should indicate the column that contains the thing you're looking for), or the thing you're looking for itself.
-   Obtain your results using the `broom` package if possible. If your model is not compatible with the broom function you're needing, then you can obtain your results by some other means, but first indicate which broom function is not compatible.

<!-------------------------- Start your work below ---------------------------->
After using the linear regression model for my two variables, I wanted to know if there was a p-value associated with the model fitting. I'm not exactly sure how this model worked with the inputted data, so I'm not sure why the p-value is 0.
```{r}
fitted_diam <- broom::tidy(model_diam)
print(fitted_diam)
```
<!----------------------------------------------------------------------------->

# Task 4: Reading and writing data

Get set up for this exercise by making a folder called `output` in the top level of your project folder / repository. You'll be saving things there.

## 4.1 (3 points)

Take a summary table that you made from Task 1, and write it as a csv file in your `output` folder. Use the `here::here()` function.

-   **Robustness criteria**: You should be able to move your Mini Project repository / project folder to some other location on your computer, or move this very Rmd file to another location within your project repository / folder, and your code should still work.
-   **Reproducibility criteria**: You should be able to delete the csv file, and remake it simply by knitting this Rmd file.

<!-------------------------- Start your work below ---------------------------->
```{r}



here::here("Users", "raveenbadyal", "MDA Raveen Badyal", "Output","exported_file.csv")
here::here()
write_csv(trees_by_neigh, here("exported_file.csv"))


```
<!----------------------------------------------------------------------------->

## 4.2 (3 points)

Write your model object from Task 3 to an R binary file (an RDS), and load it again. Be sure to save the binary file in your `output` folder. Use the functions `saveRDS()` and `readRDS()`.

-   The same robustness and reproducibility criteria as in 4.1 apply here.

<!-------------------------- Start your work below ---------------------------->

<!----------------------------------------------------------------------------->

# Overall Reproducibility/Cleanliness/Coherence Checklist 

Here are the criteria we're looking for.

## Coherence (0.5 points)

The document should read sensibly from top to bottom, with no major continuity errors. 

The README file should still satisfy the criteria from the last milestone, i.e. it has been updated to match the changes to the repository made in this milestone. 

## File and folder structure (1 points)

You should have at least three folders in the top level of your repository: one for each milestone, and one output folder. If there are any other folders, these are explained in the main README.

Each milestone document is contained in its respective folder, and nowhere else.

Every level-1 folder (that is, the ones stored in the top level, like "Milestone1" and "output") has a `README` file, explaining in a sentence or two what is in the folder, in plain language (it's enough to say something like "This folder contains the source for Milestone 1").

## Output (1 point)

All output is recent and relevant:

-   All Rmd files have been `knit`ted to their output md files. 
-   All knitted md files are viewable without errors on Github. Examples of errors: Missing plots, "Sorry about that, but we can't show files that are this big right now" messages, error messages from broken R code
-   All of these output files are up-to-date -- that is, they haven't fallen behind after the source (Rmd) files have been updated.
-   There should be no relic output files. For example, if you were knitting an Rmd to html, but then changed the output to be only a markdown file, then the html file is a relic and should be deleted.

Our recommendation: delete all output files, and re-knit each milestone's Rmd file, so that everything is up to date and relevant.

## Tagged release (0.5 point)

You've tagged a release for Milestone 2. 

### Attribution

Thanks to Victor Yuan for mostly putting this together.
